---
title: "Assignment 2"
author: "B166247"
date: "3 3 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Problem 1
## (a)

```{r}
#loading the required library
library(data.table)

#loading the files
lipids.dt <- data.table(read.delim("lipids.txt"))
classes.dt <- data.table(read.delim("lipid-classes.txt"))

#lipid classes can be found in the first column of lipids.dt
#Finding all lipids PC and assignment to new column lipid.class (42)
lipids.dt[grep("^[Pp]C", lipid.species), lipid.class := classes.dt[CE == "PC", Cholesterol.esters]]
#Finding all Cer (11)
lipids.dt[grep("^[Cc]er|CER", lipid.species), lipid.class := classes.dt[CE == "Cer", Cholesterol.esters]]
#Finding all DAG (16)
lipids.dt[grep("^[Dd]ag|DAG", lipid.species), lipid.class := classes.dt[CE == "DAG", Cholesterol.esters]]
#Finding all LPC (13)
lipids.dt[grep("^[Ll]pc|LPC", lipid.species), lipid.class := classes.dt[CE == "LPC", Cholesterol.esters]]
#Finding all LPE (3)
lipids.dt[grep("^[Ll]pe|LPE", lipid.species), lipid.class := classes.dt[CE == "LPE", Cholesterol.esters]]
#Finding all PE (25)
lipids.dt[grep("^[Pp]e|^PE", lipid.species), lipid.class := classes.dt[CE == "PE", Cholesterol.esters]]
#Finding all PS (8)
lipids.dt[grep("^[Pp]s|^PS", lipid.species), lipid.class := classes.dt[CE == "PS", Cholesterol.esters]]
#Finding all SM (0)
lipids.dt[grep("[Ss]m|SM", lipid.species)]
#Finding all TAG (147)
lipids.dt[grep("[Tt]ag|TAG", lipid.species), lipid.class := classes.dt[CE == "TAG", Cholesterol.esters]]

#it can be seen that 9 rows are still unoccupied (correspond to CE)
lipids.dt[is.na(lipid.class)]

```
After adding all classes provided in the classes.txt file, it can be seen that one class is missing and 9 rows with the abbreviation CE could not be annotated. Following the shorthand notation of lipids published by Liebisch et al. in 2013 CE corresponds to Cholesteryl ester (Liebisch et al., 2013). I added this class manually.
```{r}
#manual addition of Cholesteryl ester
lipids.dt[grep("\\b[C]e\\b|\\bCE\\b", lipid.species), lipid.class := "Cholesteryl ester"]
#test that all lipids have a class
stopifnot(!is.na(lipids.dt$lipid.class))

#count the number of entries corresponding to each class (one class missing only 8!)
lipids.dt[, .(count = .N), lipid.class]

```
## (b)
Since there are 9 lipid species in the table (including CE and exclusing SM), they are calculated as following:
```{r}
#calculate the Wald test (t-distributed)
lipids.dt[, "Wald.test(t.dist)":= round(((oddsratio-1)/se), 3)]

#calculate the degress of freedom
df <- 288-9-1

#calculate the p-value based on the t-distribution
lipids.dt[, "p.value.t" := signif(2*pt(abs(`Wald.test(t.dist)`), df, lower.tail = FALSE), 4)]

#calculate the p-value based on the normal distribution
lipids.dt[, "p.value.norm" := signif(2*pnorm(`Wald.test(t.dist)`, lower.tail = FALSE), 4)]

#checking the table
head(lipids.dt)
```
Present some evidence to justify if the normal approximation is acceptable in this instance. (ANSWER)

## (c)
```{r warning=FALSE}
holm.bonferroni <- function(results.dt, alpha){
  #copy and order the input dataframe
  order.dt <- results.dt[order(p.value.t),]
  
  #looping through all sorted entries
  for (row.idx in seq(1,nrow(order.dt))){
    #calculate the threshhold value for the row
    pcalc <- alpha/(276+1-row.idx)
    #cat(pcalc)
    #testing if the threshold is passed
    if (pcalc >= order.dt[row.idx, p.value.t]) {
      #threshold is not passed anymore and the loop breaks
      lastidx <- row.idx
      #cat(lastidx, "\n")
      #cat("This is the pcalc ", pcalc)
    } else {
      #nothting happens when the pcalc gets too large
    }
  }
  
  #returning the subset of the ordered table
  return(order.dt[1:lastidx])
}

```
## (d)
```{r warning=FALSE}
benjamini.hochberg <- function(results.dt, q){
  q <- 0.01
  #copy and order the input dataframe
  order.dt <- results.dt[order(p.value.t),]
  
  #looping through all sorted entries
  for (row.idx in seq(1,nrow(order.dt))){
    pcalc <- (row.idx/276)*q
    
    #testing if the threshold is valid
    if (pcalc > order.dt[row.idx, p.value.t]){
      #assigning the indeces that pass the threshold
      lastidx <- row.idx
      #cat(lastidx, "\n")
      #cat("This is the pcalc ", pcalc)
    } else {
      #pass as the threshold was passed
    }
    
  }
  
  #return the subset in ascending p value order
  return(order.dt[1:lastidx])
  
}

```
## (e)
```{r warning=FALSE}


```

## (f)

# Problem 2
## (a) Not sure how cross validation works. Is lambda already maximizing for AUC?
```{r warning=FALSE}
#loading the required libraries
library(caret)
library(pROC)
library(glmnet)

#load the data set
wdbc2.dt <- fread("wdbc2.csv", stringsAsFactors = TRUE)

#setting the seed
set.seed(1)

#create the data particioning
train.idx <- createDataPartition(wdbc2.dt$diagnosis, p=0.7)

#function from lab 4 to convert data tables into a matrix as expected by glmnet
prepare.glmnet <- function(data, formula=~ .) {
  ## create the design matrix to deal correctly with factor variables,
  ## without losing rows containing NAs
  old.opts <- options(na.action='na.pass')
  x <- model.matrix(formula, data)
  options(old.opts)
  ## remove the intercept column, as glmnet will add one by default
  x <- x[, -match("(Intercept)", colnames(x))]
  return(x)
}

#separate desig matrix from outcomes
ywdbc2.dt <- as.matrix(wdbc2.dt$diagnosis)
xwdbc2.dt <- prepare.glmnet(wdbc2.dt[,!"diagnosis"])

#fit ridge regression model with cross validation
fit.ridge <- cv.glmnet(xwdbc2.dt, ywdbc2.dt, family="binomial", alpha=0)
#fit regression with lasso penalty
fit.lasso <- cv.glmnet(xwdbc2.dt, ywdbc2.dt, family="binomial")
```
